---
title: "Aggregated Data & Special Cases"
author: "Maciej Nasinski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregated Data & Special Cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

This vignette covers `cat2cat_agg()` for pre-aggregated data, building mapping tables from hierarchical codes, and comparison with alternative approaches. It assumes you've read [Get Started with cat2cat](cat2cat.html).

```{r load-data}
library(cat2cat)
library(dplyr)
library(tidyr)

data(trans, package = "cat2cat")
```

## Aggregated data with `cat2cat_agg()`

When you have **pre-aggregated** data (counts or totals per category, not micro-level observations), use `cat2cat_agg()` with explicit mapping equations instead of a transition table.

### The verticals dataset

```{r agg-data}
data("verticals", package = "cat2cat")

agg_old <- verticals[verticals$v_date == "2020-04-01", ]
agg_new <- verticals[verticals$v_date == "2020-05-01", ]

agg_old
agg_new
```

The `verticals` dataset contains **aggregated** e-commerce category statistics:

- `vertical` - product category name
- `sales` - total sales value in that category
- `counts` - number of products in that category
- `v_date` - date of the snapshot

Between April and May 2020, the e-commerce platform reorganised its category structure:

| Change | Old categories | New categories |
|--------|---------------|----------------|
| Split  | Automotive    | Automotive1, Automotive2 |
| Merge  | Kids1, Kids2  | Kids |
| Rename + split | Home | Home, Supermarket |

### Equation syntax

Each mapping equation has the form:

```
OLD_SIDE  DIRECTION  NEW_SIDE
```

**Direction operators**:

- `%<%` or `<` - **backward**: replicates/splits the OLD period to match new encoding
- `%>%` or `>` - **forward**: replicates/splits the NEW period to match old encoding

**Wrapping multiple categories**: use `c(Cat1, Cat2)` to group several category names.

**Rule**: exactly one side must have a single category (one-to-many). Many-to-many equations are not allowed.

### Applying the equations

```{r agg-mapping}
agg <- cat2cat_agg(
  data = list(
    old = agg_old,
    new = agg_new,
    cat_var = "vertical",
    time_var = "v_date",
    freq_var = "counts"
  ),
  # Backward: split old "Automotive" into new codes
  Automotive %<% c(Automotive1, Automotive2),
  # Forward: split new "Kids" back into old codes
  c(Kids1, Kids2) %>% c(Kids),
  # Forward: new "Home" & "Supermarket" both map to old "Home"
  Home %>% c(Home, Supermarket)
)
```

### How proportions work

When one category splits into many, proportions come from the **target** encoding's `freq_var`:

- **Backward** (`%<%`): proportions from the NEW period counts
- **Forward** (`%>%`): proportions from the OLD period counts

For example, `Automotive %<% c(Automotive1, Automotive2)` splits old "Automotive" using the ratio of Automotive1 to Automotive2 counts in the new period.

### Variance for proportional splits

When splitting aggregates by proportion $p$, the variance of the allocated count depends on the original variance and the split:

$$\text{Var}(n \cdot p) = p^2 \cdot \text{Var}(n) + n^2 \cdot \text{Var}(p) + \text{Var}(n) \cdot \text{Var}(p)$$

If the original count $n$ is observed without error and only $p$ has sampling uncertainty, this simplifies to $\text{Var}(n \cdot p) = n^2 \cdot \text{Var}(p)$. Estimate $\text{Var}(p)$ from the target period’s category frequencies using the multinomial variance: $\text{Var}(\hat{p}_j) = \hat{p}_j(1-\hat{p}_j)/N$ where $N$ is the total count used to compute proportions.

### Inspecting the result

```{r agg-inspect}
# Old period - "Automotive" was split into two rows
agg$old[agg$old$vertical %in% c("Automotive1", "Automotive2"), ]

# New period - "Kids" was split into two rows  
agg$new[agg$new$vertical %in% c("Kids1", "Kids2"), ]
```

The `prop_c2c` column tells you how much of the original row's values to count. Rows not affected by any equation have `prop_c2c = 1`.

### Aggregating to a unified encoding

```{r agg-unified}
agg %>%
  bind_rows() %>%
  group_by(v_date, vertical) %>%
  summarise(
    sales = sum(sales * prop_c2c),
    counts = sum(counts * prop_c2c),
    .groups = "drop"
  ) %>%
  arrange(vertical, v_date)
```

Now both periods use the same category names and values are redistributed proportionally.

### When to use `cat2cat_agg()`

| Data type | Function | Example |
|-----------|----------|---------|
| Micro-data (individual observations) | `cat2cat()` | Survey responses, individual transactions |
| Aggregated data (counts/totals per category) | `cat2cat_agg()` | Published statistics, summary reports |

`cat2cat_agg()` is necessary when:

- You only have access to published summary statistics, not the underlying micro-data
- Privacy restrictions prevent access to individual-level data
- The data was collected at the aggregate level (e.g., industry reports)

## Combining micro-data and aggregated data

A common scenario: you have micro-data for some periods and only published aggregates for others. The workflow combines `cat2cat()` and `cat2cat_agg()`:

```{r combined-workflow, eval=FALSE}
# Micro-data periods: use cat2cat()
micro_result <- cat2cat(
  data = list(old = survey_2008, new = survey_2010,
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "backward")
)

# Aggregate periods: use cat2cat_agg() with same mapping logic
agg_result <- cat2cat_agg(
  data = list(old = published_2006, new = published_2008,
              cat_var = "sector", time_var = "year",
              freq_var = "employment"),
  sector_old %<% c(sector_new1, sector_new2)
)

# Combine: aggregate the micro results to match published format
micro_aggregated <- micro_result$old %>%
  group_by(g_new_c2c) %>%
  summarise(employment = sum(wei_freq_c2c * weight))

# Now both can be analyzed together at the aggregate level
```

The key insight: aggregate-level comparisons remain valid as long as you apply consistent mapping equations across periods.

## Building mapping tables from hierarchical codes

When an official transition table is **not available**, you can build one from the hierarchical structure of the codes themselves. Many classifications use a digit hierarchy: the first digit encodes the major group, the first two digits the sub-major group, and so on.

In the `trans` table, old codes are 4 digits and new codes are 6 digits. You can construct a coarser mapping by truncating to the first N digits:

```{r hierarchical-codes}
head(trans, 5)

# Build a 2-digit mapping from the full codes
trans_2digit <- data.frame(
  old = substr(trans$old, 1, 2),
  new = substr(trans$new, 1, 2)
)
trans_2digit <- unique(trans_2digit)
cat("2-digit mapping rows:", nrow(trans_2digit),
    "vs full mapping rows:", nrow(trans))
```

The coarser the grouping, the fewer many-to-many relationships (and fewer replications), but the unified categories are broader. This approach works for any classification where codes share a hierarchical prefix structure (ISCO, ICD, NACE, CPC, HS trade codes, etc.).

### Validation: filtering mapped data to coarser groups

A key property of cat2cat is that **filtering replicated data to a coarser grouping recovers the original population** (with proper weights). For complete validation examples with regression comparisons, see the [Sensitivity Analysis & Holdout Validation](cat2cat_validation.html) vignette.

Briefly: truncating codes to fewer digits creates coarser mappings with fewer many-to-many relationships. Weighted counts from mapped data at any aggregation level match original counts - confirming the replication-and-weighting procedure is neutral.

### Common hierarchical classifications

| Classification | Domain | Hierarchy |
|---------------|--------|-----------|
| ISCO | Occupations | 1-digit -> 2-digit -> 3-digit -> 4-digit |
| ICD | Diseases | Chapter -> Block -> Category -> Subcategory |
| NACE | Industries | Section -> Division -> Group -> Class |
| CPC | Products | Section -> Division -> Group -> Class -> Subclass |

## Alternative approaches

**Manual expert mapping**: A domain specialist assigns each old-to-new category pair manually. Gold standard for few, well-understood mappings; does not scale to large many-to-many tables.

**`cat2cat_agg()` for aggregated data**: When only aggregate counts are available (not micro-data), uses mapping equations to redistribute counts proportionally.

**Analyse periods separately**: Each period under its own classification. Avoids the mapping problem but precludes longitudinal modelling and trend analysis.

**When to prefer `cat2cat()`**: Many-to-many mappings, micro-data available, need a unified variable across time, and probabilistic uncertainty should be reflected in the analysis.

## Comparison with Multiple Imputation

Readers familiar with Multiple Imputation (MI) may notice a surface similarity: both create multiple versions of observations. However, **cat2cat is not MI**. MI draws independent datasets from a posterior; cat2cat uses deterministic replication from mapping tables. MI uses Rubin’s combining rules for variance; cat2cat uses a d.f. scaling correction. The methods address different problems - MI handles missing data under MAR assumptions, while cat2cat handles classification changes with known mapping structure.

## Next steps

- [Get Started with cat2cat](cat2cat.html) - basic two-period workflow
- [Multi-Period Chaining](cat2cat_multi_period.html) - chaining across 3+ time periods
- [Sensitivity Analysis & Holdout Validation](cat2cat_validation.html) - weight methods, pruning, ML validation
- [Regression on Replicated Data](cat2cat_regression.html) - SE correction with `summary_c2c()`, fixed effects models
- [Panel Data with Subject Identifiers](cat2cat_panel.html) - `id_var` for rotational panels with consistent subject IDs
