---
title: "Get Started with cat2cat"
author: "Maciej Nasinski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

## What is cat2cat?

**cat2cat** harmonizes categorical variables whose encoding changes over time - occupation codes (ISCO revisions), disease classifications (ICD-9 -> ICD-10), industry codes (NACE), product taxonomies.

| Data type | Example | Function |
|-----------|---------|----------|
| Micro-data | Survey responses, administrative records | `cat2cat()` |
| Aggregated counts | Published statistics, summary tables | `cat2cat_agg()` |
| Panel data | Subjects observed in multiple periods | `cat2cat()` with `id_var` |

## The Problem

Classifications evolve. 
When they do, **one old category maps to multiple new ones** (and vice versa):
         
```{r load-data, echo=FALSE}
library(cat2cat)
data(trans, package = "cat2cat")
```

```{r show-problem}
# Old occupation code "1111" became three different codes in the new system
trans[trans$old == "1111", ]
```

A worker coded `1111` in 2008 could be `111101`, `111102`, or `111103` in 2010. 
Which one? The original data doesn't say.

Naive approaches fail:

- Ignoring the problem leads to biased comparisons
- Manual assignment is arbitrary and not reproducible

## The Solution

**cat2cat** uses a statistically principled approach:

1. Replicate each observation onto all candidate categories  
2. Weight each replica by its probability (frequencies or ML)
3. Correct standard errors for the artificial sample inflation

Formally, for each original observation $i$ mapped to $k$ candidate categories, the weights satisfy:

$$w_{ij} \geq 0 \quad \text{and} \quad \sum_{j=1}^{k} w_{ij} = 1$$

This constraint ensures that weighted statistics equal their original values - the replication is *neutral* for the first two moments of any estimator.

For estimands that do not directly depend on the unresolved mapped category, the weighted analysis preserves unbiasedness under the stated assumptions. See [Nasinski et al. (2020)](https://doi.org/10.24425/cejeme.2020.134747) for the theoretical foundation and [Nasinski & Gajowniczek (2023)](https://doi.org/10.1016/j.softx.2023.101525) for R implementation details.

## Identification & Assumptions

You observe a worker coded "1111" in 2008 that could be "111101", "111102", or "111103" in 2010. cat2cat assigns probability weights to each candidate. The question is: are those weights correct?

The mapping table alone does not imply a single weighting - each method below adds an assumption that does. 
The choice becomes an identifying assumption for downstream estimates that depend on the mapped category (e.g. occupation fixed effects or interactions with `g_new_c2c`); estimates that do not depend on the mapped category are unaffected.
See [Regression on Replicated Data](cat2cat_regression.html#identification-and-inference) for the model-level picture.

### Three weighting schemes

Each weighting method adds one identifying assumption:

**Naive weights** (`wei_naive_c2c`): Assume uniform distribution - each candidate equally likely.
$$w_j = \frac{1}{k}$$
This is the maximum entropy (least informative) prior. It requires only the mapping table - no data from either period. Use when you have no information favoring any candidate, or as a robustness bound.

**Frequency weights** (`wei_freq_c2c`): Assume ambiguous observations distribute like the base period population.
$$w_j = \frac{n_j^{\text{base}}}{\sum_{j'} n_{j'}^{\text{base}}}$$
This requires observed counts in the base period. It works well when ambiguous cases resemble the overall population; it breaks down when they differ systematically (e.g., workers in transitional occupations may not match the stable population).

**ML weights** (`wei_*_c2c` for knn, rf, lda, nb): Assume individual features predict category membership.
$$w_j = P(j \mid X_i, g)$$
This requires training data with both the category and predictive features. ML can correct for heterogeneity that frequency weights miss - if a 25-year-old programmer is more likely in "111101" than a 55-year-old manager, ML captures this. Use `cat2cat_ml_run()` to verify ML improves over baselines.

### When each assumption matters

| Method | Assumption fails when... | Practical check |
|--------|--------------------------|-----------------|
| Naive | Distribution is highly skewed | Compare `wei_naive_c2c` vs `wei_freq_c2c` - large differences suggest uniform is wrong |
| Frequency | Ambiguous cases differ from population | Compare ML vs freq accuracy in `cat2cat_ml_run()` |
| ML | Features don't predict category | ML accuracy ≈ frequency accuracy in `cat2cat_ml_run()` |

### Reducing assumption dependence

The weighting assumption only affects observations that are replicated (`rep_c2c > 1`). Strategies to reduce replication:

- **Panel `id_var`**: returning subjects get direct matches - no replication. See [Panel Data with Subject Identifiers](cat2cat_panel.html).
- **Forward mapping**: replicates the newer (more granular) period, which often maps 1-to-1 backward.
- **Code truncation**: collapsing 6-digit to 4-digit codes creates more 1-to-1 mappings in hierarchical systems (ISCO, ICD, NACE).

If most observations have `rep_c2c = 1`, results depend minimally on the assumption.

**Sensitivity check**: run analyses with `wei_naive_c2c`, `wei_freq_c2c`, and ML weights. Similar conclusions -> robust. Divergent -> report the range.

## When cat2cat won't help

**Blockers** (cat2cat cannot be applied):

- **No mapping table**: cat2cat requires a crosswalk. Alternative: expert coding or separate analyses per period.
- **Unobserved category**: cat2cat harmonizes *observed* codes. Missing category is a missing-data problem - impute first.

**Method-specific limits** (workarounds exist):

- **Empty base categories**: frequency weights fall back to naive (1/k). Use `wei_naive_c2c` or ML with external training data.
- **Few observations per ML class**: models fail or overfit. Use frequency weights or coarser categories.

Naive weights (`wei_naive_c2c`) always work - they need only the mapping table.

**Diagnostic**: check `mean(result$old$rep_c2c)` and `cor(wei_freq_c2c, wei_naive_c2c)`. Very high replication or correlation near 1 suggests the mapping is too diffuse to improve over uniform weights. See [Multi-Period Chaining](cat2cat_multi_period.html#reducing-replication-via-truncation) for truncation strategies.

## Use Cases

**Repeated cross-sections** (most common): harmonize independent surveys using base-period frequencies. The Quick Example below demonstrates this.

**Multi-period analysis** (3+ waves): chain `cat2cat()` calls sequentially - each step uses the previous step's mapped frequencies. See [Multi-Period Chaining](cat2cat_multi_period.html).

**Panel data** (subjects observed in multiple periods): pass `id_var` to directly match returning subjects - no replication for them, only new entrants go through the probabilistic path. See [Panel Data with Subject Identifiers](cat2cat_panel.html).

**Aggregated data** (published counts only, no micro-data): use `cat2cat_agg()` with explicit equations:

```{r use-case-agg}
data(verticals, package = "cat2cat")

agg_old <- verticals[verticals$v_date == "2020-04-01", ]
agg_new <- verticals[verticals$v_date == "2020-05-01", ]

agg_result <- cat2cat_agg(
  data = list(
    old = agg_old,
    new = agg_new,
    cat_var = "vertical",
    time_var = "v_date",
    freq_var = "counts"
  ),
  # Backward mapping: old Automotive split into Automotive1 + Automotive2
  Automotive %<% c(Automotive1, Automotive2),
  # Forward mapping: Kids1 + Kids2 merged into Kids  
  c(Kids1, Kids2) %>% c(Kids)
)

agg_result$old[c("vertical", "prop_c2c", "counts")]
```

See [Aggregated Data & Special Cases](cat2cat_aggregated.html) for equation syntax.

## Key Concepts

### Direction: Backward vs Forward

| Direction | Base period | Replicated period | Result encoding |
|-----------|-------------|-------------------|-----------------|
| `"backward"` | NEW | OLD | New (modern) codes |
| `"forward"` | OLD | NEW | Old (legacy) codes |

```{r direction-comparison}
# Setup for comparison
occup_2008 <- occup[occup$year == 2008, ]
occup_2010 <- occup[occup$year == 2010, ]

# Backward: old period gets replicated onto new codes
backward <- cat2cat(
  data = list(old = occup_2008, new = occup_2010, 
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "backward")
)

# Forward: new period gets replicated onto old codes
forward <- cat2cat(
  data = list(old = occup_2008, new = occup_2010, 
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "forward")
)

# Which period gets replicated depends on direction
cat("Backward: old period replicated from", nrow(occup_2008), "to", nrow(backward$old), "rows\n")
cat("Forward: new period replicated from", nrow(occup_2010), "to", nrow(forward$new), "rows")
```

### Weights

| Weight | Source | Use when... |
|--------|--------|-------------|
| `wei_freq_c2c` | Base period frequencies | Default choice |
| `wei_naive_c2c` | Uniform (1/k) | Robustness checks |
| `wei_knn_c2c` | k-Nearest Neighbours on features | Non-linear boundaries, no distributional assumption |
| `wei_rf_c2c` | Random Forest on features | Feature interactions, larger training sets |
| `wei_lda_c2c` | Linear Discriminant Analysis | Fast, assumes normality & equal covariance |
| `wei_nb_c2c` | Naive Bayes (via `e1071`) | Mixed feature types, small samples |

ML weights are added only when you pass an `ml` argument to `cat2cat()`. Validate them against naive and frequency baselines with `cat2cat_ml_run()` - see [Sensitivity Analysis & Holdout Validation](cat2cat_validation.html).

## Quick Example

```{r quick-example}
data(occup, package = "cat2cat")

occup_2008 <- occup[occup$year == 2008, ]
occup_2010 <- occup[occup$year == 2010, ]

result <- cat2cat(
  data = list(
    old = occup_2008,
    new = occup_2010,
    cat_var = "code",
    time_var = "year"
  ),
  mappings = list(trans = trans, direction = "backward")
)
```

**What happened?** 2008 observations were replicated onto 2010 category codes. One worker may appear multiple times with different `g_new_c2c` values:

```{r show-replication}
# A replicated observation (rep_c2c > 1 means replicated)
result$old[result$old$rep_c2c > 1, ][1:3, 
           c("code", "g_new_c2c", "wei_freq_c2c", "rep_c2c")]
```

### Sanity Check: Weights Preserve Totals

Weights must sum to 1 per original observation. This ensures weighted statistics equal original statistics:

```{r sanity-check}
# Row counts
c(original = nrow(occup_2008), 
  replicated = nrow(result$old),
  weighted = sum(result$old$wei_freq_c2c))

# Mean salary preserved
c(original = mean(occup_2008$salary),
  weighted = weighted.mean(result$old$salary, result$old$wei_freq_c2c))
```

### Naive vs Frequency Weights

Both weight types are always available. Naive weights (`wei_naive_c2c`) assign equal probability to each candidate - useful as a robustness baseline:

```{r naive-vs-freq}
# Compare weights for a replicated observation
result$old[result$old$rep_c2c > 1, ][1:3, 
           c("g_new_c2c", "wei_naive_c2c", "wei_freq_c2c")]

# Same analysis with naive weights (robustness check)
c(freq_mean = weighted.mean(result$old$salary, result$old$wei_freq_c2c),
  naive_mean = weighted.mean(result$old$salary, result$old$wei_naive_c2c))
```

If results are similar, conclusions don't depend on the distributional assumption. Large differences warrant investigation.

### Standard Error Correction
 
Replication inflates row count -> artificially small standard errors. The correction formula scales standard errors by the replication factor:

$$\text{SE}_{\text{corrected}} = \text{SE}_{\text{naive}} \times \sqrt{\frac{n_{\text{rep}}}{n_{\text{orig}}}}$$

P-values are recomputed from $t(n_{\text{orig}})$ distribution. Use `summary_c2c()` for regression:

```{r se-correction}
model <- lm(log(salary) ~ age + sex + edu, 
            data = result$old, weights = wei_freq_c2c)

# Corrected SEs (uses original df)
summary_c2c(model, df_old = nrow(occup_2008))
```

## Diagnostic Plot

```{r plot, fig.width=7, fig.height=3.5, fig.alt="Diagnostic plot displaying weight histograms and replication statistics for mapped observations"}
plot_c2c(result$old, type = "both")
```

## Learn More

| Topic | Vignette |
|-------|----------|
| Multi-period chaining (3+ waves) | [Multi-Period Chaining](cat2cat_multi_period.html) |
| Sensitivity analysis, ML validation | [Sensitivity Analysis & Holdout Validation](cat2cat_validation.html) |
| SE correction theory, fixed effects models | [Regression on Replicated Data](cat2cat_regression.html) |
| Panel data assumptions, `id_var` details | [Panel Data with Subject Identifiers](cat2cat_panel.html) |
| Aggregated counts, equation syntax | [Aggregated Data & Special Cases](cat2cat_aggregated.html) |
