---
title: "Get Started"
author: "Maciej Nasinski"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(size = "tiny")
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
```

## `cat2cat`

The introduced `cat2cat` algorithm was designed to offer an easy and clear interface to apply a transition table which was provided by a data maintainer or built by a researcher. The objective is to unify an inconsistent coded categorical variable in a panel dataset, where a transition table is the core element of the process.

Examples of a dataset with such inconsistent coded categorical variable are ISCO (The International Standard Classification of Occupations) or ICD (International Classification of Diseases) based one. The both classifications are regularly updated to adjust to e.g. new science achievements. More clearly we might image that e.g. new science achievements brings new occupations types on the market or enable recognition of new diseases types.

The categorical variable encoding changes are typically provided by datasets providers in the form of transition table, for each time point the changes occurred. A transition table conveys information needed for matching all categories between two periods of time. More precisely it contains two columns where the first column contains old categories and the second column contains the new ones.
Sometimes a transition table has to be created manually by a researcher.

The main rule is to replicate the observation if it could be assigned to a few categories.
More precisely for each observation we look across a transition table to check how the original category could be mapped to the opposite period one. Then using simple frequencies or statistical methods to approximate probabilities of being assigned to each of them.
For each observation that was replicated, the probabilities have to add up to one.
The algorithm distinguishes different mechanics for panel data with and without unique identifiers.

## Core elements

There should be highlighted 3 important elements:

1. Type of the data - panel dataset with unique identifiers vs panel dataset without unique identifiers, aggregate data vs unit data.
2. Transition table. 
3. Direction of a transition, forward or backward - a new or an old encoding as a base one.

## Manual transitions
### Aggregate dataset

```{r}
library(cat2cat)
library(dplyr)

data(verticals)
agg_old <- verticals[verticals$v_date == "2020-04-01", ]
agg_new <- verticals[verticals$v_date == "2020-05-01", ]

## cat2cat_agg - could map in both directions at once although 
## usually we want to have old or new representation

agg <- cat2cat_agg(data = list(old = agg_old, 
                              new = agg_new, 
                              cat_var = "vertical", 
                              time_var = "v_date",
                              freq_var = "counts"), 
                  Automotive %<% c(Automotive1, Automotive2),
                  c(Kids1, Kids2) %>% c(Kids),
                  Home %>% c(Home, Supermarket))
            
## possible processing
  
agg$old %>% 
group_by(vertical) %>% 
summarise(sales = sum(sales*prop_c2c), counts = sum(counts*prop_c2c), v_date = first(v_date))

agg$new %>% 
group_by(vertical) %>%
summarise(sales = sum(sales*prop_c2c), counts = sum(counts*prop_c2c), v_date = first(v_date))
```

## Automatic using trans table

### Dataset with unique identifiers

```{r}
## the ean variable is a unique identifier
data(verticals2)

vert_old <- verticals2[verticals2$v_date == "2020-04-01", ]
vert_new <- verticals2[verticals2$v_date == "2020-05-01", ]

## get transitions table
trans_v <- vert_old %>% 
inner_join(vert_new, by = "ean") %>%
select(vertical.x, vertical.y) %>% distinct()

# 
## cat2cat
## it is important to set id_var as then we merging categories 1 to 1 
## for this identifier which exists in both periods.
verts <- cat2cat(
  data = list(old = vert_old, new = vert_new, id_var = "ean", cat_var = "vertical", time_var = "v_date"),
  mappings = list(trans = trans_v, direction = "backward")
)
```

### Dataset without unique identifiers

```{r}
data(occup)
data(trans)

occup_old <- occup[occup$year == 2008,]
occup_new <- occup[occup$year == 2010,]

## cat2cat
occup_simple <- cat2cat(
  data = list(old = occup_old, new = occup_new, cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "backward")
)

## with informative features it might be usefull to run ml algorithm
## currently only knn, lda or rf (randomForest),  a few methods could be specified at once 
## where probability will be assessed as fraction of closest points.
occup_2 <- cat2cat(
  data = list(old = occup_old, new = occup_new, cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "backward"),
  ml = list(method = "knn", features = c("age", "sex", "edu", "exp", "parttime", "salary"), 
            args = list(k = 10))
)

# summary_plot
plot_c2c(occup_2$old, type = c("both"))

# mix of methods
occup_2_mix <- cat2cat(
  data = list(old = occup_old, new = occup_new, cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "backward"),
  ml = list(method = c("knn", "rf", "lda"), features = c("age", "sex", "edu", "exp", "parttime", "salary"), 
            args = list(k = 10, ntree = 50))
)
# correlation between ml models and simple fequencies
occup_2_mix$old %>% select(wei_knn_c2c, wei_rf_c2c, wei_lda_c2c, wei_freq_c2c) %>% cor()
# cross all methods and subset one highest probability category for each subject
occup_old_mix_highest1 <- occup_2_mix$old %>% 
                cross_c2c(.) %>% 
                prune_c2c(.,column = "wei_cross_c2c", method = "highest1") 
```

## Regression

The replication process is neutral for calculating at least the first 2 central moments for all variables.
This is because for each observation which was replicated, probabilities sum to one.
If we are removing non-zero probability observations then replication probabilities have to be reweighed to still sum to one.
Important note is that removing non zero probability observations should be done only if needed, as it impact the counts of categorical variable levels. More preciously removing non-zero weights will influence the regression model if we will use the unified categorical variable.

### Regression

The next 3 regressions have the same results.

```{r}
## orginal dataset 
lms2 <- lm(I(log(salary)) ~ age + sex + factor(edu) + parttime + exp, occup_old, weights = multiplier)
summary(lms2)

## using one highest cross weights
## cross_c2c to cross differen methods weights
## prune_c2c - highest1 leave only one the highest probability obs for each subject
occup_old_2 <- occup_2$old %>% 
                cross_c2c(., c("wei_freq_c2c", "wei_knn_c2c"), c(1/2,1/2)) %>% 
                prune_c2c(.,column = "wei_cross_c2c", method = "highest1") 
lms <- lm(I(log(salary)) ~ age + sex + factor(edu) + parttime + exp, occup_old_2, weights = multiplier)
summary(lms)

## we have to adjust size of stds as we artificialy enlarge degrees of freedom
occup_old_3 <- occup_2$old %>% 
                prune_c2c(method = "nonzero") #many prune methods like highest
lms_replicated <- lm(I(log(salary)) ~ age + sex + factor(edu) + parttime + exp, occup_old_3, weights = multiplier * wei_freq_c2c)
# Adjusted R2 is meaningless here
lms_replicated$df.residual <- nrow(occup_old) - length(lms_replicated$assign)
suppressWarnings(summary(lms_replicated))
```

### Regression with unified variable

Example regression model with usage of the unified variable (`g_new_c2c`).
A separate model for each occupational group.

```{r}
occup_old_4 <- occup_2$old %>% 
                prune_c2c(method = "nonzero") #many prune methods like highest
formula_oo <- formula(I(log(salary)) ~ age + sex + factor(edu) + parttime + exp + factor(year))
oo <- rbind(occup_old_4, occup_2$new) %>% 
  group_by(g_new_c2c) %>% 
  do(
    lm = tryCatch(
      summary(lm(formula_oo, ., weights = multiplier * wei_freq_c2c)), 
      error = function(e) NULL
    )
  ) %>%
  filter(!is.null(lm))

head(oo)

oo$lm[[1]]
```

