---
title: "Multi-Period Chaining"
author: "Maciej Nasinski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Period Chaining}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

This vignette assumes you've read [Get Started](cat2cat.html) and understand the basic `cat2cat()` workflow. 
Here we tackle chaining across multiple periods.

```{r load-data}
library(cat2cat)
library(dplyr)

data(occup, package = "cat2cat")
data(trans, package = "cat2cat")

occup_2006 <- occup[occup$year == 2006, ]
occup_2008 <- occup[occup$year == 2008, ]
occup_2010 <- occup[occup$year == 2010, ]
occup_2012 <- occup[occup$year == 2012, ]
```

## Mapping table stability analysis

Before diving into multi-period chaining, it's useful to understand how "disruptive" your mapping table is. 
Hierarchical codes (like ISCO occupations) often preserve higher-level groupings even when detailed codes change.

```{r trans-stability}
# What percentage of mappings have different codes at each digit level?
max_digits <- max(nchar(as.character(trans[[1]])), nchar(as.character(trans[[2]])))

stability <- sapply(1:max_digits, function(d) {
  old_trunc <- substr(as.character(trans[[1]]), 1, d)
  new_trunc <- substr(as.character(trans[[2]]), 1, d)
  mean(old_trunc != new_trunc) * 100
})

data.frame(
  digits = 1:max_digits,
  pct_changed = round(stability, 1)
)
```

**Interpretation:**

- **Low % at 1-2 digits**: Broad categories are stable; most disruption is in detailed codes
- **High % at all levels**: Major restructuring occurred; expect more replication and higher sensitivity to weight choice
- **Increasing %**: Typical hierarchical pattern - fine-grained codes change more than coarse ones

This informs your analysis strategy:

- Stable mappings -> results likely robust across weight methods
- Disruptive mappings -> sensitivity analysis is critical; consider aggregating to coarser levels

### Reducing replication via truncation

Truncating codes to fewer digits can substantially reduce replication. This comparison shows backward and forward mapping at full resolution (4->6 digit) vs truncated (4->4 digit):

```{r truncation-diagnostic}
# Truncate codes and mapping table to 4 digits
occup_2008_trunc <- occup_2008
occup_2008_trunc$code <- substr(occup_2008_trunc$code, 1, 4)
occup_2010_trunc <- occup_2010
occup_2010_trunc$code <- substr(occup_2010_trunc$code, 1, 4)
trans_trunc <- unique(data.frame(
  old = substr(trans$old, 1, 4),
  new = substr(trans$new, 1, 4)
))

# Backward: full resolution (4->6) vs truncated (4->4)
back_full <- cat2cat(
  data = list(old = occup_2008, new = occup_2010, 
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "backward")
)
back_trunc <- cat2cat(
  data = list(old = occup_2008_trunc, new = occup_2010_trunc,
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans_trunc, direction = "backward")
)

# Forward: full resolution vs truncated
fwd_full <- cat2cat(
  data = list(old = occup_2008, new = occup_2010, 
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "forward")
)
fwd_trunc <- cat2cat(
  data = list(old = occup_2008_trunc, new = occup_2010_trunc,
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans_trunc, direction = "forward")
)

# Compare replication and weight agreement
data.frame(
  mapping = c("backward full (4->6)", "backward trunc (4->4)",
              "forward full (6->4)", "forward trunc (4->4)"),
  mean_rep = c(mean(back_full$old$rep_c2c), mean(back_trunc$old$rep_c2c),
               mean(fwd_full$new$rep_c2c), mean(fwd_trunc$new$rep_c2c)),
  cor_freq_naive = c(cor(back_full$old$wei_freq_c2c, back_full$old$wei_naive_c2c),
                     cor(back_trunc$old$wei_freq_c2c, back_trunc$old$wei_naive_c2c),
                     cor(fwd_full$new$wei_freq_c2c, fwd_full$new$wei_naive_c2c),
                     cor(fwd_trunc$new$wei_freq_c2c, fwd_trunc$new$wei_naive_c2c))
)
```

Forward mapping at full resolution often has the lowest replication because many detailed new codes map 1-to-1 to coarser old codes. Truncating both sides to 4 digits creates some many-to-many mappings that weren't present before. Backward mapping typically benefits most from truncation - the printed table above shows the reduction for this dataset.

## Chaining across multiple periods

Real panels have more than two waves. The `occup` data has four (2006, 2008, 2010, 2012) and the encoding changed between 2008 and 2010. To bring all waves onto the same encoding, apply `cat2cat()` iteratively - one pair of adjacent periods at a time - feeding the output of one step into the next.

> **Error propagation warning**: Each chaining step introduces mapping uncertainty. Weights from step 1 become frequency inputs for step 2, so errors accumulate across steps. Long chains warrant sensitivity analysis at each link, and consider whether coarser category groupings would reduce cascading uncertainty.

### Backward: unifying to 2010+ codes

We start from the transition point and work backwards. Step 1 maps 2008 onto 2010 codes. Step 2 maps 2006 onto the already-mapped 2008.

```{r backward-chain}
# Step 1: 2008 -> 2010
step1 <- cat2cat(
  data = list(old = occup_2008, new = occup_2010,
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "backward")
)

# Frequencies from step 1 give us more control in step 2
freqs_2008 <- step1$old %>%
  group_by(g_new_c2c) %>%
  summarise(counts = round(sum(wei_freq_c2c)))

# Step 2: 2006 -> mapped-2008
# The 2008 data now uses g_new_c2c as its category variable
step2 <- cat2cat(
  data = list(old = occup_2006, new = step1$old,
              cat_var_old = "code", cat_var_new = "g_new_c2c",
              time_var = "year"),
  mappings = list(trans = trans, direction = "backward",
                  freqs_df = freqs_2008)
)
```

Now assemble the four periods. Periods that were never replicated (2010, 2012) need `dummy_c2c()` to get the standard `_c2c` columns:

```{r assemble-backward-panel}
panel_back <- bind_rows(
  step2$old,                       # 2006 - replicated in step 2
  step1$old,                       # 2008 - replicated in step 1
  step1$new,                       # 2010 - base period, not replicated
  dummy_c2c(occup_2012, "code")    # 2012 - untouched
)
```

Validation - the weighted count in each year should match the original:

```{r validate-backward-counts}
panel_back %>%
  group_by(year) %>%
  summarise(weighted_n = round(sum(wei_freq_c2c))) %>%
  left_join(count(occup, year), by = "year")
```

### Forward: unifying to pre-2010 codes

Forward mapping replicates the newer period. Codes that exist in the new data but not in the transition table need a placeholder:

```{r forward-chain}
trans_fwd <- rbind(
  trans,
  data.frame(old = "no_cat",
             new = setdiff(c(occup_2010$code, occup_2012$code), trans$new))
)

# Step 1: 2008 -> 2010 (forward replicates 2010)
fwd1 <- cat2cat(
  data = list(old = occup_2008, new = occup_2010,
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans_fwd, direction = "forward")
)

freqs_2010 <- fwd1$new %>%
  group_by(g_new_c2c) %>%
  summarise(counts = round(sum(wei_freq_c2c)))

# Step 2: mapped-2010 -> 2012 (forward replicates 2012)
fwd2 <- cat2cat(
  data = list(old = fwd1$new, new = occup_2012,
              cat_var_old = "g_new_c2c", cat_var_new = "code",
              time_var = "year"),
  mappings = list(trans = trans_fwd, direction = "forward",
                  freqs_df = freqs_2010)
)

panel_fwd <- bind_rows(
  dummy_c2c(occup_2006, "code"),  # untouched
  fwd1$old,                        # 2008 - base
  fwd1$new,                        # 2010 - replicated in step 1
  fwd2$new                         # 2012 - replicated in step 2
)
```

### Adding ML to the chain

ML weights integrate naturally into this workflow. Define the ML setup once and reuse it:

```{r ml-chain}
ml_setup <- list(
  data = bind_rows(occup_2010, occup_2012),
  cat_var = "code",
  method = "knn",
  features = c("age", "sex", "edu", "exp", "parttime", "salary"),
  args = list(k = 10)
)

step1_ml <- cat2cat(
  data = list(old = occup_2008, new = occup_2010,
              cat_var = "code", time_var = "year"),
  mappings = list(trans = trans, direction = "backward"),
  ml = ml_setup
)

step2_ml <- cat2cat(
  data = list(old = occup_2006, new = step1_ml$old,
              cat_var_old = "code", cat_var_new = "g_new_c2c",
              time_var = "year"),
  mappings = list(trans = trans, direction = "backward"),
  ml = ml_setup
)

panel_back_ml <- bind_rows(
  step2_ml$old,
  step1_ml$old,
  step1_ml$new,
  dummy_c2c(occup_2012, "code", ml = "knn")
)
```

After chaining, prune per year to control the growing row count:

> **Computational note**: Each chaining step has complexity $O(k \times n)$ where $k$ is the average number of candidate categories per mapping and $n$ is the number of observations. For large datasets with complex mappings, memory usage can grow substantially - consider pruning between steps or processing in batches by coarse category groups.

```{r prune-chain}
panel_pruned <- panel_back_ml %>%
  split(.$year) %>%
  lapply(function(x) {
    x %>% cross_c2c() %>%
      prune_c2c(column = "wei_cross_c2c", method = "highest1")
  }) %>%
  bind_rows()

all.equal(nrow(panel_pruned), nrow(occup))
```
