---
title: "Panel Data with Subject Identifiers"
author: "Maciej Nasinski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Panel Data with Subject Identifiers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

This vignette covers the `id_var` feature for panel data where subjects have consistent identifiers across periods. It assumes you've read [Get Started with cat2cat](cat2cat.html).

```{r load-data}
library(cat2cat)
library(dplyr)

data("occup_panel", package = "cat2cat")
data("trans", package = "cat2cat")
```

## The problem: rotational panels

Most real-world labour force panels are **rotational**: not all workers appear in every wave. Some drop out (attrition), new workers enter (refreshment), and only a subset appears in consecutive waves.

When the occupational encoding changes between waves, you face two types of workers:

1. **Returning workers** - observed in both periods. If we trust their identifier, we know their exact category mapping (old -> new) without uncertainty.
2. **New/exiting workers** - observed in only one period. These must go through the standard mapping-table replication.

The `id_var` argument tells `cat2cat()` to exploit returning workers for direct matching, reducing unnecessary replication.

## The occup_panel dataset

The `occup_panel` dataset is a simulated rotational panel derived from `occup`. Unlike the main `occup` dataset (repeated cross-sections where each worker appears once), `occup_panel` includes workers who return across waves:

```{r panel-structure}
# Panel overview
cat("Total observations:", nrow(occup_panel), "\n")
cat("Unique workers (panel_id):", n_distinct(occup_panel$panel_id), "\n\n")

cat("Observations per year:\n")
print(table(occup_panel$year))

cat("\nObservations per quarter:\n")
print(table(occup_panel$quarter))

cat("\nHow many waves does each worker appear in?\n")
appearances <- occup_panel %>% count(panel_id) %>% count(n, name = "workers")
print(appearances)
```

The critical feature: workers who span the 2009Q4-2010Q1 encoding change boundary:

```{r panel-encoding-change}
# Use quarters adjacent to encoding change (2009Q4 = old encoding, 2010Q1 = new encoding)
panel_old <- occup_panel[occup_panel$quarter == "2009Q4", ]
panel_new <- occup_panel[occup_panel$quarter == "2010Q1", ]

shared_ids <- intersect(panel_old$panel_id, panel_new$panel_id)
cat("Workers observed in both 2009Q4 AND 2010Q1:", length(shared_ids), "\n")
```

These workers provide direct evidence of how their old (4-digit) codes map to new (6-digit) codes - information we can exploit with `id_var`.

## Mapping with `id_var`

```{r idvar-mapping}
result <- cat2cat(
  data = list(
    old = panel_old,
    new = panel_new,
    id_var = "panel_id",
    cat_var = "code",
    time_var = "quarter"
  ),
  mappings = list(trans = trans, direction = "backward")
)
```

## How `id_var` works

When `id_var` is specified, `cat2cat` splits the target-period data into two groups:

1. **Direct match**: workers whose `panel_id` appears in *both* periods. Category is mapped 1-to-1 from what we observe in the base period - no replication. They receive `wei_freq_c2c = 1` and `rep_c2c = 1`.
2. **Replication path**: workers appearing *only* in the target period. These go through the standard mapping-table replication.

```{r idvar-inspect}
# Check replication counts
cat("Replication distribution:\n")
print(table(result$old$rep_c2c))

# Direct-match workers have rep_c2c = 1
direct_match <- result$old %>% filter(rep_c2c == 1)
cat("\nDirect matches (returning workers):", nrow(direct_match), "\n")

# Replicated workers have rep_c2c > 1
replicated <- result$old %>% filter(rep_c2c > 1)
cat("Replicated rows (old-period-only workers):", nrow(replicated), "\n")
cat("Unique old-period-only workers:", n_distinct(replicated$index_c2c), "\n")
```

The weighted total still equals the original count:

```{r idvar-validate}
cat("Weighted sum:", sum(result$old$wei_freq_c2c), "\n")
cat("Original count:", nrow(panel_old), "\n")
```

## Inspecting direct matches

For returning workers, we can verify their code mapping:

```{r direct-match-inspect}
# A returning worker's trajectory
example_worker <- shared_ids[1]

# Their 2009Q4 observation (before mapping)
panel_old %>% 
  filter(panel_id == example_worker) %>%
  select(panel_id, quarter, code, age, salary)

# Their 2010Q1 observation (in base period)
panel_new %>%
  filter(panel_id == example_worker) %>%
  select(panel_id, quarter, code, age, salary)

# After mapping (in result$old)
result$old %>%
  filter(index_c2c == which(panel_old$panel_id == example_worker)) %>%
  select(panel_id, quarter, code, g_new_c2c, wei_freq_c2c, rep_c2c)
```

The returning worker's `g_new_c2c` is directly taken from their 2010 observation - no uncertainty, no replication.

## The core assumption

> **Direct matching assumes a worker's true occupation does not change between adjacent waves.**

For `occup_panel`, this means: if worker #123 was in "1111" (old code) in 2009Q4 and "111101" (new code) in 2010Q1, we trust that mapping. The worker didn't genuinely change occupations - the *encoding* changed.

This assumption is **reasonable** when:

- **Short inter-wave gaps** - gaps of 1–3 years are typically short for major career changes in most labor markets (though this varies by industry and demographics)
- **Stable occupations** - most workers don't change jobs between waves
- **Reliable identifiers** - `panel_id` truly tracks the same person

This assumption is **suspect** when:

- **Long panels** - people change careers over 10+ years
- **High mobility** - young workers, gig economy, frequent job changes
- **ID issues** - identifiers recycled or subject to error
- **Attrition bias** - if workers who leave the sample differ systematically from those who stay, direct matching on survivors may not generalize to the full population

## Testing the assumption

Before trusting `id_var` matching, compare returning workers to single-period workers. If they differ systematically, the direct-match assumption may not hold:

```{r test-assumption}
# Identify returning vs single-period workers in old period
returning_ids <- intersect(panel_old$panel_id, panel_new$panel_id)

panel_old_test <- panel_old %>%
  mutate(is_returning = panel_id %in% returning_ids)

# Compare characteristics
panel_old_test %>%
  group_by(is_returning) %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    mean_salary = mean(salary, na.rm = TRUE),
    pct_female = mean(sex == "female", na.rm = TRUE) * 100,
    mean_exp = mean(exp, na.rm = TRUE)
  ) %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))
```

**Interpretation:**

- **Similar characteristics**: Returning workers are representative; `id_var` matching is appropriate.
- **Systematic differences**: Returning workers differ (e.g., older, more experienced, higher paid). Direct matching may work for *them*, but replication-based mapping may be more appropriate for the full sample. Consider running analyses with and without `id_var` as sensitivity check.

### Attrition and cat2cat

When subjects attrit (leave the panel), they cannot be directly matched in later periods. cat2cat handles this naturally - attrited subjects from the old period go through replication-based mapping, while returning subjects get direct matches via `id_var`. If attrition is non-random (e.g., correlated with occupation), consider standard attrition corrections (inverse probability weighting) *before* passing data to cat2cat.

## When to use `id_var`

**Decision guide:**

1. **Do subjects have consistent identifiers across periods?**
   - NO: `id_var` not applicable - use standard replication
   - YES: Continue to step 2

2. **Is the inter-wave gap short relative to category mobility?**
   - NO (many waves apart, or mobility is high): Consider omitting `id_var` - category changes may reflect real movement
   - YES: Continue to step 3

3. **Is category mobility low in your population?**
   - YES: **Use `id_var`** - direct matches reduce unnecessary replication
   - NO (high mobility): **Consider omitting `id_var`** - observed changes may be real, not encoding artifacts

| Scenario | Recommendation |
|----------|---------------|
| Short-term panel, stable occupations | **Use `id_var`** - genuine changes unlikely |
| Long panel with substantial career mobility | **Consider omitting `id_var`** - changes may be real |
| Product catalogs with EAN/SKU codes | **Use `id_var`** - products don't change identity |
| Repeated cross-sections | `id_var` **not applicable** - no shared subjects |

## Comparing with and without `id_var`

What happens if we ignore the identifiers?

```{r idvar-compare}
# WITHOUT id_var - all old-period workers go through replication
result_no_id <- cat2cat(
  data = list(
    old = panel_old,
    new = panel_new,
    cat_var = "code",
    time_var = "quarter"
  ),
  mappings = list(trans = trans, direction = "backward")
)

cat("WITH id_var:\n")
cat("  Rows:", nrow(result$old), "\n")
cat("  Workers with rep_c2c = 1:", sum(result$old$rep_c2c == 1), "\n")
cat("  Average replication:", round(mean(result$old$rep_c2c), 2), "\n")

cat("\nWITHOUT id_var:\n")
cat("  Rows:", nrow(result_no_id$old), "\n")
cat("  Workers with rep_c2c = 1:", sum(result_no_id$old$rep_c2c == 1), "\n")
cat("  Average replication:", round(mean(result_no_id$old$rep_c2c), 2), "\n")
```

Without `id_var`, returning workers are unnecessarily replicated - introducing artificial uncertainty for cases where we have direct evidence.

## Regression comparison

Does `id_var` affect regression results? For non-mapped variables, coefficients should be similar (both are unbiased), but standard errors may differ due to different replication levels:

```{r idvar-regression}
# Simple wage regression
f <- I(log(salary)) ~ age + sex + factor(edu) + exp + parttime

# With id_var
model_with <- lm(f, data = result$old, 
                 weights = multiplier * wei_freq_c2c)

# Without id_var
model_without <- lm(f, data = result_no_id$old, 
                    weights = multiplier * wei_freq_c2c)

# Compare coefficients
coef_compare <- data.frame(
  with_id = coef(model_with),
  without_id = coef(model_without)
)
coef_compare$diff <- coef_compare$with_id - coef_compare$without_id
round(coef_compare, 4)
```

> Replication inflates the row count, so naive standard errors are optimistic. Adjust them with `summary_c2c(model, df_old = nrow(panel_old))` - see [Regression on Replicated Data](cat2cat_regression.html) for details.

```{r idvar-regression-se}
summary_c2c(model_with, df_old = nrow(panel_old))
```

Coefficients are similar (both unbiased), but using `id_var` reduces unnecessary replication, keeps the data closer to the original structure, and leverages direct observation rather than probabilistic inference for returning workers.

## Example 2: E-commerce product panels

The `verticals2` dataset demonstrates `id_var` with product EAN codes in an e-commerce context:

```{r verticals-example}
data("verticals2", package = "cat2cat")

vert_old <- verticals2[verticals2$v_date == "2020-04-01", ]
vert_new <- verticals2[verticals2$v_date == "2020-05-01", ]

# Build transition table from overlapping products
trans_v <- vert_old %>%
  inner_join(vert_new, by = "ean") %>%
  select(vertical.x, vertical.y) %>%
  distinct()

# Map with id_var
verts <- cat2cat(
  data = list(
    old = vert_old, new = vert_new,
    id_var = "ean",
    cat_var = "vertical",
    time_var = "v_date"
  ),
  mappings = list(trans = trans_v, direction = "backward")
)

cat("Products with direct match:", sum(verts$old$rep_c2c == 1), "\n")
cat("Products requiring replication:", n_distinct(verts$old$index_c2c[verts$old$rep_c2c > 1]), "\n")
```

**Why `id_var` is ideal for products:** Unlike workers who might change occupations, products don't change their fundamental identity - an EAN code uniquely identifies the same product across time. When category reorganizations occur (e.g., "Electronics" split into "Computers" and "Mobile Devices"), directly observed product category assignments provide ground truth for the mapping.

## Next steps

- [Get Started with cat2cat](cat2cat.html) - basic two-period workflow
- [Multi-Period Chaining](cat2cat_multi_period.html) - chaining across 3+ time periods
- [Sensitivity Analysis & Holdout Validation](cat2cat_validation.html) - weight methods, pruning, ML validation
- [Regression on Replicated Data](cat2cat_regression.html) - SE correction with `summary_c2c()`, fixed effects
- [Aggregated Data & Special Cases](cat2cat_aggregated.html) - `cat2cat_agg()` for pre-aggregated data
